Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-19 ms.tex
fixed another sum
MWF changes

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $w_k$ such that $w_k \ne w_j$ for $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a diagonal positive matrix with diagonal elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $M^n_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{w_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

This model is a special case of the general model (Eq.~\ref{eq:model}) where
\begin{equation} 
{M} = \frac{1}{n-1} \begin{bmatrix}
0 & 1 & 1 & \ldots & 1 \\
1 & 0 & 1 & \ldots & \vdots \\
1 & 1 & 0 & \ldots & \vdots \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \ldots & \ldots & \ldots & 0
\end{bmatrix},
\end{equation}
with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
Note that here ${M}$ is irreducible.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots > w_n$),
and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
If columns of ${M}$ are identical
\begin{equation}
{M} = \begin{bmatrix}
m_1 & m_1 & \ldots & m_1 \\
m_2 & m_2 & \ldots & m_2 \\
\vdots & \vdots & \vdots & \vdots \\
m_n & m_n & \ldots & m_n
\end{bmatrix},
\end{equation}
with $\sum_{k=1}^{n}{m_k}=1$,
then $m_k$ can be considered the relative population size of deme $A_k$ --
this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.

Similarly, if demes are arranged in a circle, for example around a lake, 
then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
and ${M}$ takes the form
\begin{equation}
{M} = \begin{bmatrix}
p_0 & p_1 & \ldots & p_{n-1} \\
p_{n-1} & p_0 & \ldots & p_{n-2} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_2 & \ldots & p_0
\end{bmatrix},
\end{equation}
where $p_k \ge 0$ and $\sum_{k=1}^{n}{p_k}=1$.

@@

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppD}; and
(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.

The following lemma presents a \emph{mean fitness principle} for the
sensitivity of the stable population mean fitness $\bar{w}^*$ to changes
in $C_k$, the probability of transition from $A_k$.

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
\end{lemma}

\begin{proof}
$\bar{w}^*$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{v_k} = 1$ and $\sum_k{u_k v_k} = 1$.

Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
\end{equation}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v = 
u\tr (0 - Z_k + M Z_k)D v = \\
-v_k u_k w_k + v_k w_k u\tr[M]_k = \\
v_k w_k (u\tr[M]_k - u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
u\tr \bar{w}^* = u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\bar{w}^*$ and the $k$
element of $u$:
\begin{equation}
u_k \bar{w}^* = (1-C_k) w_k u_k + C_k w_k u\tr [M]_k.
\end{equation}
Multiplying both sides by $v_k$ (which is positive if $w_k>0$) and rearranging, we get 
\begin{equation}
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k) = v_k w_k (u\tr [M]_k - u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - w_k).
\end{equation}
Finally, note that $u_k, v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (\bar{w}^* - w_k),
\end{equation}
which completes the proof.
\end{proof}

The above lemma provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then increasing $C_k$ will increase the population mean fitness.

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k w_k (u\tr[M]_k - u_k) = \\
w_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

Given $0 < w_k \le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - w_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k w_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

% Modifier model
We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
We will show that modifier alleles that increase the stable population mean fitness in accordance with Lemma~\ref{lemma:mfp} are favored by natural selection.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}). 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Lemma~\ref{lemma:mfp}).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).
Note that, in general, $\bar{w}^*$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - w_k).
\end{equation}

@@

\begin{proof}
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium} and \nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}\big\rvert_{\tilde{C}_k = C_k}$ is $\lambda_1=1$.

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$, and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^n)_{j,k} > 0$ for all $j,k$ for some natural $n$.
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness.

@@

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = \big(1 - C_k\big) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{j \ne k}{C_j \frac{w_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1, 2, \ldots, n$, where $\bar{w} = \sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

This model is a special case of the general model (Eq.~\ref{eq:model}) where
\begin{equation} 
{M} = \frac{1}{n-1} \begin{bmatrix}
0 & 1 & 1 & \ldots & 1 \\
1 & 0 & 1 & \ldots & \vdots \\
1 & 1 & 0 & \ldots & \vdots \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \ldots & \ldots & \ldots & 0
\end{bmatrix},
\end{equation}
with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
Note that here ${M}$ is irreducible and primitive.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $w_k$ ($w_0 \ge w_1 \ge \ldots \ge w_n$),
and the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

@@

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main diagonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible and primitive.

@@

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where matrix $M$ is primitive, $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
If columns of ${M}$ are identical
\begin{equation}
{M} = \begin{bmatrix}
m_1 & m_1 & \ldots & m_1 \\
m_2 & m_2 & \ldots & m_2 \\
\vdots & \vdots & \vdots & \vdots \\
m_n & m_n & \ldots & m_n
\end{bmatrix},
\end{equation}
with $m_k>0$ and $\sum_{k=1}^{n}{m_k}=1$,
then $m_k$ can be considered the relative population size of deme $A_k$ --
this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.

Similarly, if demes are arranged in a circle, for example around a lake, 
then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
and ${M}$ takes the form
\begin{equation}
{M} = \begin{bmatrix}
p_0 & p_1 & \ldots & p_{n-1} \\
p_{n-1} & p_0 & \ldots & p_{n-2} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_2 & \ldots & p_0
\end{bmatrix},
\end{equation}
where $p_k > 0$ and $\sum_{k=1}^{n}{p_k}=1$.

@@

Write the equilibrium  frequency vector $f$ in Eq.~\ref{eq:model} as $\hat v$ and the stable population mean fitness as $\hat{\bar w}$, then
\begin{equation}\label{eq:model_equilibrium}
\hat{\bar w} \hat v = (I-C+MC)D \hat v.
\end{equation}
Note that (i) the existence and uniqueness of $\hat{\bar w}$ and $\hat v$ are guaranteed by the \emph{Perron-Frobenius theorem}~(\cite{Otto2007}) because $(I-C+MC)D$ is a non-negative primitive matrix; 
(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppC}.

The following result presents a \emph{mean fitness principle} for the
sensitivity of the equilbrium mean fitness $\hat{\bar w}$ to changes
in $C_k$, the probability of transition from $A_k$.
\medskip

\noindent{\bf Result 1. Mean fitness principle.}
{\sl Let $\hat{\bar w}$ be the leading eigenvalue of $(I-C+MC)D$,
and $\hat u$ and $\hat v$ be the corresponding left and right eigenvectors.
Then,
\begin{equation}\label{mfp}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \hat{\bar w}}{\partial C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.}

\begin{proof}
$\hat{\bar w}$ is the leading (real positive) eigenvalue of $(I-C+MC)D$ (see eq.~\ref{eq:model_equilibrium}),
and the corresponding left and right eigenvectors $u$ and $v$ are both non-negative and uniquely defined by $\sum_k{\hat v_k} = 1$ and $\sum_k{\hat u_k \hat v_k} = 1$.

Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppA}),
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v.
\end{equation}

@@

Then,
\begin{equation}\label{mfp_intermediate}
\begin{aligned}
\hat u\tr \frac{\partial (I-C+MC)D}{\partial C_k} \hat v = 
\hat u\tr (0 - Z_k + M Z_k)D \hat v = \\
-\hat v_k \hat u_k w_k + \hat v_k w_k \hat u\tr[M]_k = \\
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k). 
\end{aligned}
\end{equation}

The corresponding equation to Eq.~\ref{eq:model_equilibrium} for the left
eigenvector $u$ is 
\begin{equation}
\hat u\tr \hat{\bar w} = \hat u\tr (I - C + MC) D,
\end{equation}
which gives us a relation between $\hat{\bar w}$ and the $k$
element of $u$:
\begin{equation}
\hat u_k \hat{\bar w} = (1-C_k) w_k \hat u_k + C_k w_k \hat u\tr [M]_k.
\end{equation}
Multiplying both sides by $\hat v_k$  and rearranging, we get 
\begin{equation}
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k) = \hat v_k w_k (\hat u\tr [M]_k - \hat u_k),
\end{equation}
which when substituted into Eq.~\ref{mfp_intermediate} yields:
\begin{equation}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\frac{\hat u_k \hat v_k}{C_k} (\hat{\bar w} - w_k).
\end{equation}
Finally, note that $\hat u_k, \hat v_k, C_k \ge 0$ and therefore
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat{\bar w} - w_k),
\end{equation}
which completes the proof.
\end{proof}

The above result provides a condition for the effect of changing $C_k$, the probability for transition from $A_k$, on the stable population mean fitness.
Specifically, if $A_k$ individuals have below-average fitness, then increasing $C_k$ will increase the population mean fitness.

@@

\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial C_k} = 
\hat v_k w_k (\hat u\tr[M]_k - \hat u_k) = \\
w_k (\sum_{j=1}^{n}{\hat u_j M_{j,k} \hat v_k} - \hat u_k \hat v_k).
\end{aligned}
\end{equation}
Here, $\hat u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppB}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $\hat u_k \hat v_k$ is the \emph{ancestor frequency} of type $k$~\citep{Hermisson2002}, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{\hat u_i M_{i,k} \hat v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occurring.

Given $0 < w_k\le 1$, from Eq.~\ref{mfp_intermediate} we have the following corollary.

\begin{corollary}[Reproductive value principle]
In the notation of Result 1,
\begin{equation}
sign \frac{\partial \hat{\bar w}}{\partial C_k} = 
sign (\hat u\tr [M]_k - \hat u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

@@

We turn our attention to the case where the transition occurs from a subset $K$ of the types.
For example, \citet[Appendix~A]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, 
and a modifier allele that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
If the transition rates are controlled by a threshold rule, then 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases}.
\end{equation}
If $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types in $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \hat{\bar w}}{\partial \tau} = 
\hat u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) \hat v = \\
\frac{1}{\tau} \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)}.
\end{aligned}
\end{equation}
Thus
\begin{multline}\label{mfp_sign_threshold}
sign \frac{\partial \hat{\bar w}}{\partial \tau} = 
sign \sum_{k \in K}{\hat u_k \hat v_k (\hat{\bar w} - w_k)} = \\ 
sign \Big(\hat{\bar w} - \frac{\sum_{k \in K}{\hat u_k \hat v_k w_k}}{\sum_{k \in K}{\hat u_k \hat v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{mfp_sign}.
Therefore, increased transition from types in $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types in $K$ is below the stable population mean fitness.

@@

% Modifier model
We now focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine $C_k$, the rates of transition from the different types.
We will show that modifier alleles that increase the stable population mean fitness in accordance with Result 1 are favored by natural selection.

@@

The frequencies in the next generation, $f'$ for allele $m$ and $g'$ for allele $M$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Here, $\bar{w}=\sum_{k=1}^{n}{(f_k + g_k) w_k}$ is the mean fitness of the entire population. 
Note that Eq.~\ref{eq:model} is the special case of Eq.~\ref{eq:modifier_model} where allele $M$ is absent; i.e. $g_k=0$ for all $k$.

@@

We have found a condition for a modifier allele that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness (Result 1).
Could such an allele increase in frequency when initially rare in the population?
To answer this we analyze the stability of resident modifier allele $m$ with transition rates $C_k$ to the invasion by a modifier allele $M$ with rates $\tilde{C}_k$, see Eq.~\ref{eq:modifier_model}~\citep[for an introduction to \emph{evolutionary invasion analysis}, see][ch.~12]{Otto2007}

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(v,0)$ in the space $\sum_{k=1}^n{\hat v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
\end{equation}
where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

@@

\begin{proof}
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\hat{\bar w}$,
\begin{equation}
\hat{\bar w} \cl_{ex}\big\rvert_{\tilde{C}_k = C_k} = (I-C+MC) D,
\end{equation}
and since $\hat{\bar w}$ is the leading eigenvalue of the RHS (see Eq.~\ref{eq:model_equilibrium}),
the leading eigenvalue of $\cl_{ex}\big\rvert_{\tilde{C}_k = C_k}$ is $\lambda_1=1$.

Now, applying Result 1 (eq.~\ref{mfp_sign}) to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda_1$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k} =
sign\Big(\lambda_1 - \frac{w_k}{\hat{\bar w}}\Big).
\end{equation}
Thus
\begin{multline}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign\Big(1 - \frac{w_k}{\hat{\bar w}}\Big) =  
sign(\hat{\bar w} - w_k),
\end{multline}
since $\hat{\bar w}>0$.
This completes the proof.
\end{proof}
